---
title: "Vaccination Rates and COVID mortality in Chicago"
author: William F Parker, MD, MS
output:
  html_document:
    theme: darkly
--- 

```{r global_options, include=FALSE, cache = FALSE}
library(knitr)
opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, linewidth=60, cache= TRUE)
library(tidyverse)
library(ggpubr)
library(sf)
library(tigris)
library(tidycensus)
options(tigris_use_cache = TRUE)
```

```{r}
#zip_codes <- zctas(cb = TRUE, class = "sf")

#save(file = "zip_code_data.R", zip_codes)

load("zip_code_data.R")

zips_sf <- zip_codes %>%
  select(zip = ZCTA5CE10, geometry)
```

```{r}
vax_data_by_zip <- read_csv("https://data.cityofchicago.org/api/views/553k-3xzc/rows.csv?accessType=DOWNLOAD") %>%
  mutate(zip = `Zip Code`) %>% 
  select(zip, `1st Dose - Percent Population`) %>%
  group_by(zip) %>%
  summarise(first_dose_rate = max(`1st Dose - Percent Population`)) %>%
  mutate(pct_vax = first_dose_rate)
```



```{r}
covid_by_zip <- read_csv("https://data.cityofchicago.org/api/views/yhhz-zm2v/rows.csv?accessType=DOWNLOAD") %>%   
  mutate(zip = as.character(`ZIP Code`))%>%
  group_by(zip) %>% 
  summarise(total_cases = sum(`Cases - Weekly`, na.rm =  TRUE ),
            case_rate = max(`Case Rate - Cumulative`,  na.rm =  TRUE),
              death_rate = max(`Death Rate - Cumulative`),
            total_deaths = sum(`Deaths - Weekly`),
            population = mean(Population)
  ) %>% mutate(death_rate_check = 100000*total_deaths/population)
```


```{r}
# covid_by_zip %>%
#   arrange(death_rate) %>%
#   select(zip, death_rate, total_deaths) %>% 
#   filter(zip %in% c(60613, 60657, 60610))
```


```{r}
# covid_by_zip %>%
#   arrange(death_rate) %>%
#   select(zip, death_rate, total_deaths) %>% 
#   filter(zip %in% c(60649))
```


```{r}
to_map <- vax_data_by_zip %>% 
  left_join(covid_by_zip %>% select(zip, death_rate)) %>%
  select(zip, pct_vax, death_rate) %>% 
  filter(is.infinite(pct_vax) == FALSE & zip != "Unknown")
```


```{r}
plot_A <- zips_sf %>%
  left_join(to_map) %>% 
  filter(is.na(pct_vax) == FALSE) %>%
  mutate(pct_vax = cut(100*pct_vax, breaks = seq(0,30,5))) %>%
  ggplot(aes(geometry = geometry, fill = pct_vax)) + 
  geom_sf() + 
  theme_void() + coord_sf() + 
  #scale_fill_distiller(, limits = c(0, 100*max(to_map$pct_vax))) + 
  #scale_fill_viridis_c(direction = -1) +
  labs(fill = "1st dose (%)", title = "Vaccination") +
    theme(legend.position = "right", 
          plot.title = element_text(hjust = 0.5),
          legend.title = element_text(size =16)) +
    scale_fill_brewer(direction = 1)

```

```{r}
plot_B <- zips_sf %>%
  left_join(to_map) %>% 
  filter(is.na(pct_vax) == FALSE) %>%
  mutate(death_rate = cut(death_rate, 5)) %>%
  ggplot(aes(geometry = geometry, fill = death_rate)) + 
  geom_sf() + 
  theme_void() + coord_sf() + 
  #scale_fill_distiller(direction = 1, limits = c(0, max(to_map$death_rate)), palette = "Reds") + 
  #scale_fill_viridis_c(option = "inferno") +
  labs(fill = "Mortality Rate\n(Deaths/100,000)", title = "Deaths")+
    theme(legend.position = "right", plot.title = element_text(hjust = 0.5)) +
    scale_fill_brewer(palette = "Reds", direction = 1) 
```

# City of Chicago Data Sources

* [Vaccine source data](https://data.cityofchicago.org/Health-Human-Services/COVID-19-Vaccine-Doses-by-ZIP-Code-1st-Dose/c28u-q29v) 
* [Mortality source data](https://data.cityofchicago.org/Health-Human-Services/COVID-19-Cases-Tests-and-Deaths-by-ZIP-Code/yhhz-zm2v)

Last updated at `r format(Sys.Date(),'%A, %B %d')` at `r format(Sys.time(), '%I:%M %p')` CST.  

Code is available [here](https://github.com/08wparker/CDPH_vaccine_allocation.git).

# Map
```{r}
ggarrange(plot_B, plot_A, ncol = 2, nrow = 1)

ggsave("vax_death_map_chicago.pdf")
```

Mortality rates are cumulative COVID-19 deaths to date per 100,000 residents of the zip code. Vaccination rates are percentage of population in the zip code receiving the **first dose**. 

# Correlation
```{r}
to_map %>%
  ggplot(aes(x = death_rate, y = 100*pct_vax))+
  geom_point() + labs(x = "Deaths per 100,000 residents", y = "Residents vaccinated (%)")
```

The correlation between morality and vaccination is `r cor(to_map$pct_vax, to_map$death_rate) %>% format(digits =2 )`







